---
title: "Script Gr 02 (NEU!)"
output: html_document
---

Installing different packages needed for the following code and loading the library
```{r}
library(ggplot2)
library(gridExtra)
library(reshape2)
library(data.table)
library(cluster)
library(rstudioapi)
library(pheatmap)
library(caret)
library(tidyverse)
library(dendextend)
library(factoextra)
library(devtools)
library(ggfortify)
library(rstudioapi)  
library(data.table) 
library(scales) 
library(stats)
library(caTools)
```

Setting the sys-path
```{r}
root.dir = dirname(rstudioapi::getSourceEditorContext()$path)
```


Part 1: Data Cleanup

Extracting and splitting our data. 
The mutation data is different from the other matrixes, so we define a new matrix only containing the mtation data
```{r}
data = readRDS(paste0(root.dir,"/DepMap19Q1_allData.RDS"))
mut <- data$mutation 
```

Additionally to the mutation matrix we want another matrix only containing the data which is not the mutation matrix.
We define which data should be excluded
```{r}
'%!in%' <- function(x,y)!('%in%'(x,y)) 
dt_new <- lapply(which(names(data) %!in% "mutation"), function(a) data[[a]]) 
names(dt_new) <- names(data)[which(names(data) %!in% "mutation")] 
```

defining the samples so we can extract only our cancer type of interest
```{r}
sample_case = c("Skin Cancer") 
``` 

We look at the annotation matrix an search only for the Primary diseases which match the previous defined sample case (Skin cancer) 
```{r}
samples = data$annotation$DepMap_ID[which(data$annotation$Primary.Disease == sample_case)]
```

extracting only the celllines with skin cancer as the primary disease from our matrices (dt_new) and save the new matrices in a new large list (processed_data)
```{r}
processed_data <- lapply(1:length(dt_new), function(a) {  
  dat_picker <- dt_new[[a]]  
  output <- dat_picker[,which(colnames(dat_picker) %in% samples)]
  output <- output[complete.cases(output),]
  output <- output[order(rownames(output)),]
  return(output)
})
names(processed_data) <- names(dt_new)

processed_data$annotation <- data$annotation[which(data$annotation$DepMap_ID %in% samples),]
```

Now we extract our cellines from the mutation data
```{r}
ids = which(names(mut) %in% samples) 
allDepMap_mutation_SkinCancer = lapply(ids, function(a) {
  mut[[a]]})
```

Losing the mutations which are not deleterious meaning not interesting to us
```{r}
allDepMap_mutation_SkinCancer = lapply(1:34, function(a) {allDepMap_mutation_SkinCancer[[a]][which(allDepMap_mutation_SkinCancer[[a]][,"isDeleterious"]== TRUE), ]})
```

losing all the genes which are not in every data frame 
```{r}
 a <- unique(c(rownames(processed_data[[1]]),rownames(processed_data[[2]]),rownames(processed_data[[3]]),rownames(processed_data[[4]])))

i <- 1
out <- vector("character", length(seq_along(1:16970)))
for (x in seq_along(a)) {
  if(a[x] %in% rownames(processed_data$expression) & a[x] %in% rownames(processed_data$copynumber) & a[x] %in% rownames(processed_data$kd.ceres) & a[x] %in% rownames(processed_data$kd.prob))
  {out[i] <- a[x]
  i <- i+1
  } 
}

processed_data2 <- processed_data

processed_data2 <- lapply(processed_data, function(a) {
  a <- a[which(rownames(a) %in% out),]
  return(a)
})
processed_data2$annotation <- processed_data$annotation
processed_data <- processed_data2
rm(processed_data2)
```

Part 2: Data vizualisation

Preparing our data for plotting

Creating new data for plotting out of our data of interest
```{r}
finalPlottingData <- lapply(1:(length(processed_data)-1), function(a) {
  dtPicker <- processed_data[[a]]
  out <- melt(dtPicker) 
  out$Gene <- rep(rownames(dtPicker), ncol(dtPicker)) 
  out$Case <- names(processed_data)[1:(length(processed_data)-1)][a] 
  colnames(out) <- c("Sample", "Value", "Gene", "Case") 
  return(out)
})
names(finalPlottingData) <- names(processed_data)[1:(length(processed_data)-1)] 
lapply(finalPlottingData, head)
```

Plotting the mutation data
```{r}
singleGenes <- as.vector(unique(as.data.frame(rbindlist(lapply(seq_along(allDepMap_mutation_SkinCancer), function(a) {out <- as.data.frame(as.vector(unique(allDepMap_mutation_SkinCancer[[a]]$Hugo_Symbol)))}))))[,1])


geneCounts <- sapply(seq_along(singleGenes), function(a) {
  genePicker <- singleGenes[a] 
  print(paste0("I am doing: ", a))
  sumGene <- lapply(seq_along(allDepMap_mutation_SkinCancer), function(b) {
    mutPicker <- allDepMap_mutation_SkinCancer[[b]] 
    out <- as.data.frame(length(which(mutPicker$Hugo_Symbol == genePicker))) 
    return(out)
  })
  geneCount <- colSums(as.data.frame(rbindlist(sumGene))) 
  return(geneCount)
})
names(geneCounts) <- singleGenes 
geneCounts <- as.data.frame(geneCounts) 
colnames(geneCounts) <- c("Value")
sortedGenCounts <- geneCounts[order(-geneCounts$Value), , drop = FALSE]
head(sortedGenCounts) 
```

Extact the data for the top 40
```{r}
plotData <- sortedGenCounts[1:10, ,drop = FALSE]
topDriverGenes <- rownames(plotData)

dataTopDriverGenes <- lapply(1:length(processed_data), function(a) { 
  dat_picker <- processed_data[[a]] 
  output <- dat_picker[which(rownames(dat_picker) %in% topDriverGenes),]
  return(output)
})
names(dataTopDriverGenes) <- names(dt_new)
lapply(dataTopDriverGenes, head)
```


Extrac the data for epression and ceres
```{r}
driverexpression <- dataTopDriverGenes$expression
driverkd.ceres <- dataTopDriverGenes$kd.ceres
```

putting the names of the matrixes in one allDep_mutation 
```{r}
names(allDepMap_mutation_SkinCancer) <- samples

OneMatrix <- data.frame()
for (i in c(1:34)) {
  OneMatrix <- rbind(OneMatrix, allDepMap_mutation_SkinCancer[[i]][,Hugo_Symbol:DepMap_ID])
}

ZelllinesMutations <- OneMatrix[which(OneMatrix$Hugo_Symbol %in% topDriverGenes ),]
ZelllinesMutations <- cbind(ZelllinesMutations$Hugo_Symbol, ZelllinesMutations$DepMap_ID)
```

creating a heatmap with the kd.ceres values -> not the whole matrix just the first ten genes and all the celllines, 
#take care blue means really, really significant p-value!
```{r}
pheatmap(as.matrix(processed_data$kd.ceres[25:45, 1:34]), clustering_method = "ward.D2",border_color = "white", fontsize = 10, 
         main = paste0("kdCERES for potential 2nd site targets"),
         show_rownames = F, show_colnames = T,
         cutree_rows = 4,
         cutree_cols = 2, 
         fontsize_row=10)  
```
# -> the knockout of the different genes influences the cells in different ways

Creating boxplots
```{r}
ggplot(data = finalPlottingData$expression, aes(x=Sample, y=Value)) +
  geom_boxplot(aes(fill = Sample), outlier.size = 0.1, outlier.alpha = 0.2) + 
  theme_bw(base_size = 7) + 
  theme(legend.position= "none", 
        legend.direction="horizontal", 
        plot.title = element_text(hjust = 0.5), 
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1), 
        legend.title= element_blank(), 
        axis.title.x = element_blank(), 
        strip.text.y = element_text(angle = 90)) 
```

Plotting the spread in the number of Mutations per Gene in a Boxplot
```{r}
geneCounts <- cbind(geneCounts, "Mutations per Gene")

ggplot(data = geneCounts, aes(x="Mutations per Gene", y=Value)) +
  geom_boxplot(aes(fill = "Mutations per Gene"), outlier.size = 2, outlier.alpha = 0.2) + 
  theme_bw(base_size = 7) + 
  theme(legend.position= "none", 
        legend.direction="horizontal", 
        plot.title = element_text(hjust = 0.5), 
        axis.text.x = element_text(angle = 0, vjust = 0.5, hjust= 0.5, size = 10), 
        legend.title= element_blank(), 
        axis.title.x = element_blank(), 
        strip.text.y = element_text(angle = 90)) 
```
# (both boxplots) -> overview over our processed data -> outliers in every cellline, there are differences between the celllines which can be examined
#boxplot mutation data nur wenige gene die viel mutieren weshalb es gut funktionieren sollte driver mutation zu definieren -> wenige die oft mutiert sind, stechen stark hervor

Plotting only the top 50 genes
```{r}
plotData <- sortedGenCounts[1:10, ,drop = FALSE]

plotData$Gene <- rownames(plotData)

ggplot(data = plotData) +
  (geom_bar(mapping = aes(x = Gene, y = Value), stat = "identity")) +
  theme_bw(base_size = 7) + 
  theme(legend.position= "none", 
        legend.direction="horizontal", 
        plot.title = element_text(hjust = 0.5), 
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, size = 10), 
        axis.text.y = element_text(angle = 90, vjust = 0.5, hjust=1, size = 10), 
        legend.title= element_blank(), 
        axis.title.x = element_blank(), 
        strip.text.y = element_text(angle = 90)) 
```


Part 3: Dimensionality reduction 

Hierachical cluster 

Creating a hierachical cluster with the celllines of the kd.ceres
```{r}
drivergene <- 4 
dataset <- processed_data$kd.ceres 
realcelllinenames <- processed_data$copynumber 

colnames(dataset)[which(colnames(dataset) %in% unique(ZelllinesMutations[which(ZelllinesMutations[,1] == topDriverGenes[drivergene]),2]))] <- topDriverGenes[drivergene]

cor.mat = cor(processed_data$kd.ceres[1:50,], method = "spearman")
cor.dist = as.dist(1 - cor.mat)
cor.hc = hclust(cor.dist, method = "ward.D2")
cor.hc = as.dendrogram(cor.hc)
plot(cor.hc, las = 2, cex.lab = 0.7)

colnames(dataset) <- realcelllinenames
rm(drivergene, realcelllinenames, dataset)
```
# -> we expected the driver mutatuins clustering together but they do not


k-means 
```{r}
km = kmeans(x =t(driverexpression), centers = 6, nstart = 30)
wss = sapply(1:11, function(k) {
  kmeans(x = t(driverexpression), centers = k)$tot.withinss
})
plot(1:11, wss, type = "b", pch = 19, xlab = "Number of clusters K", ylab = "Total within-clusters sum of squares")

D = dist(t(driverexpression))
plot(silhouette(km$cluster, D))


avg_sil <- function(k) {
  km.res <- kmeans(t(driverexpression), centers = k, nstart = 25)
  ss <- silhouette(km.res$cluster, dist(t(driverexpression)))
  mean(ss[, 3])
}

fviz_nbclust(t(driverexpression), kmeans, method = "silhouette")

km2 <- kmeans(t(driverexpression), centers = 2, nstart = 25)
km3 <- kmeans(t(driverexpression), centers = 6, nstart = 25)
km4 <- kmeans(t(driverexpression), centers = 4, nstart = 25)
km5 <- kmeans(t(driverexpression), centers = 9, nstart = 25)

p1 <- fviz_cluster(km2, geom = "text", labelsize = 8, data = t(driverexpression)) + ggtitle("k = 2")
p2 <- fviz_cluster(km3, geom = "text", labelsize = 8, data = t(driverexpression)) + ggtitle("k = 6")
p3 <- fviz_cluster(km4, geom = "text", labelsize = 8, data = t(driverexpression)) + ggtitle("k = 4")
p4 <- fviz_cluster(km5, geom = "text", labelsize = 8, data = t(driverexpression)) + ggtitle("k = 9")


grid.arrange(p1, p2, p3, p4, nrow = 2)

rm(km,km2,km3,km4,km5,p1,p2,p3,p4)
```
# ->clustert sich in zwei cluster nach der silhouette methode


PCA
```{r}
drivergene <- 4 # determines which of the Drivermutations will be seen in the cluster at the x axis
dataset <- processed_data$expression # determines which dataset we use
realcelllinenames <- processed_data$copynumber # is just defined so we can switch the colnames back so that they are showing the celllines after the clustering

colnames(dataset)[which(colnames(dataset) %in% unique(ZelllinesMutations[which(ZelllinesMutations[,1] == topDriverGenes[drivergene]),2]))] <- topDriverGenes[drivergene]


pca = prcomp(t(dataset), center = F, scale. = F)
summary(pca)
#zum Anzeigen von labels (Zelllinien)
autoplot(prcomp(t(dataset)), data = t(dataset), colour = 'blue', label = TRUE, label.size = 3)
str(pca)
plot(pca, type = "l")

colnames(dataset) <- realcelllinenames
rm(drivergene, realcelllinenames, dataset)
```
## Ziel insgesamt dass sich die gene zusammen clustern -> je nach ausgang aussage treffen


Part 4 statistical test
#only use the TOP 10 driver genes, #pick one driver gene, #the kdCERES matrix is of interest take its' rownames as refrence, #pick a potetnial 2nd site target, #pick the driver mut data, #pick the 2nd site data, #make a spearman correlation, #return the p-values, #rename all, #get a nice dataframe, #rename the list of lists, #look at the nice data
```{r}
driverGenes <- topDriverGenes[1:10] 
ttestgenes <- rownames(driverkd.ceres)

potSecondSites <- lapply(seq_along(driverGenes), function(a) {
  genePicker <- driverGenes[a] 
  print(paste0("I am doing driver mut: ", a))
  output <- sapply(seq_along(rownames(processed_data$kd.ceres)), function(b) { 
    secondSitePicker <- rownames(processed_data$kd.ceres)[b] 
    if (secondSitePicker != genePicker) {
      drMUT <- processed_data$kd.ceres[which(rownames(processed_data$kd.ceres) == genePicker),] 
      sndMUT <- as.vector(processed_data$kd.ceres[which(rownames(processed_data$kd.ceres) == secondSitePicker),]) 
      cor.val <- cor.test(unlist(drMUT, use.names=FALSE) , unlist(sndMUT, use.names=FALSE), method = "spearman") 
      return(cor.val$p.value) 
    } else {
      return(1)
    }
  })
  names(output) <- rownames(processed_data$kd.ceres) 
  output <- as.data.frame(output) 
  return(output)
})
names(potSecondSites) <- driverGenes 
lapply(potSecondSites, head) 
```

Ordering the data according to their p-values and creating a new data set with the potentional Target genes (Top ten genes with the lowest so most important p-value)
```{r}
potSecondSites <- lapply(potSecondSites, function(a){
  a <- as.data.frame(cbind(a$output, rownames(a)))
  a <- a[order(a[1]), ]
})
```

Selecting the 20 Genes out of every DriverGene List with the lowest p score
```{r}
potSecondSitestop20 <- lapply(seq_along(potSecondSites), function (a){
  output <- potSecondSites[[a]][1:20,]
  return(output)
})
names(potSecondSitestop20) <- driverGenes

ggplot(data = melt(potSecondSitestop20)) +
  (geom_bar(mapping = aes(x = V2, y = V1), stat = "identity")) +
  theme_bw(base_size = 7) + 
  theme(legend.position= "none", 
        legend.direction="horizontal", 
        plot.title = element_text(hjust = 0.5), 
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, size = 10), 
        axis.text.y = element_text(angle = 90, vjust = 0.5, hjust=1, size = 10), 
        legend.title= element_blank(), 
        axis.title.x = element_blank(), 
        strip.text.y = element_text(angle = 90)) 
```


Part 5: Multiple linear Regression

data
```{r}
a <- finalPlottingData$expression[,1:3]
a <-a[,c(1,3,2)]
copynumber <- finalPlottingData$copynumber[,2]
kd.ceres <- finalPlottingData$kd.ceres[,2]
kd.prob <- finalPlottingData$kd.prob[,2]

MRegressionanalysis <- cbind(a,copynumber,kd.ceres,kd.prob)

DriverexpressionTP53 <- c()

for (i in 1:34) {
  a <- 16970*i
  c <- (16970* (i-1))+1
  b <- samples[i]
  DriverexpressionTP53[c:a] <- processed_data$expression["TP53",b]
}

MRegressionanalysis <- cbind(MRegressionanalysis,DriverexpressionTP53)
```

Splitting the dataset into the Training set and Test set
```{r}
set.seed(123) 
split = sample.split(MRegressionanalysis$DriverexpressionTP53, SplitRatio = 0.8)
training_set = subset(MRegressionanalysis, split == TRUE) 
test_set = subset(MRegressionanalysis, split == FALSE) 
```


Fitting Multiple Linear Regression to the Training set
```{r}
regressor = lm(formula = DriverexpressionTP53 ~ .,
               data = training_set)
```

Predicting the Test set results
```{r}
y_pred = predict(regressor, newdata = test_set) 
test_set$Prediction = y_pred
test_set 


cor.test(test_set$Profit, test_set$Prediction, method = "spearman")



sessionInfo() 
```
#finally done:)